\section{Related Work}

It is worth comparing our theoretical results to a different algorithm for (unweighted) projected counting \cite{fichte2018exploiting}, which runs on a formula $\varphi$ in time $2^{2^{O(k)}} \cdot p(\varphi)$, where $k$ is the \emph{primal treewidth} \cite{SS10} of $\varphi$, and $p$ scales polynomially in the size of $\varphi$. Assuming the Exponential-Time Hypothesis \cite{impagliazzo2001problems}, all FPT algorithms parameterized by primal treewidth must be double-exponential \cite{fichte2018exploiting}. 
On the other hand, by Theorem \ref{thm_td_to_join} and Theorem \ref{thm:graded_from_virtual} our algorithm, based on graded project-join trees, runs in time $2^{O(k')}$, where $k'$ is the primal treewidth of $\psi$ (which we call the \emph{$\{X,Y\}$-graded primal treewidth} of $\phi$). While $k'$ is larger than $k$, on many benchmarks $k'$ is significantly smaller than $2^k$. 

Our proposed graded project-join trees can be seen as a specialization of \emph{structure trees} \cite{stearns2002exploiting} to the case of projected model counting. Sterns and Hunt \cite{stearns2002exploiting} suggest constructing structure trees by manually modifying tree decomposers to consider only structure trees respecting the variable quantification order (i.e., to consider gradedness directly). In this work, we take a different approach by using existing tools for standard project-join trees (in particular, tree decomposers) in a black-box way. This is crucial for the practical success of our tool, as we can leverage continual progress in tree decomposition. 

Projected model counting is also a special case of the \emph{functional aggregate query (FAQ)} problem \cite{KNR16}. 	Our graded project-join trees can be seen as a specialization of FAQ variable orders. \textbf{Theorem 7.5} of \cite{KNR16} gives an algorithm for constructing an FAQ variable orders from a sequence of tree decompositions, which, in the context of projected model counting, is equivalent to the technique we discussed in Section \ref{sec:planning:free} of ignoring relevant variables while planning to project irrelevant variables. In contrast, our approach can potentially find lower-width graded project-join trees by incorporating relevant variables even when planning to project irrelevant variables. It may be possible to lift this improvement to the FAQ framework in future work.

A recent framework for projected counting is \nest{} \cite{hecher2020taming}, a hybrid solver.
Similar to our framework, \nest{} includes a planning phase (using the tree-decomposition tool \htd{} \cite{AMW17}) and an execution phase (using the database engine \tool{Postgres} \cite{stonebraker1986design} and the projected counter \projmc{} \cite{lagniez2019recursive}, alongside other tools).
We predict that \nest{} may benefit from switching the projected-counting component to \procount{}, which was often faster than \projmc{} in Experiment 3.
While we were unable to run a full experimental comparison%
\footnote{\nest{} is an unweighted tool, but the benchmarks in Section \ref{sec:procount:experiments} are weighted.
Moreover, the cluster used in Section \ref{sec:procount:experiments} does not support database management systems.}
with \nest{}, we evaluated \nest{} against \procount{} on \wapsBenchmarks{} benchmarks \cite{gupta2019waps} (with weights removed) using a single CPU core of an Intel i7-7700HQ processor (2.80-GHz) with 30 GB of RAM.
\procount{} and \nest{} respectively solved 69 and 59 benchmarks, with a 100-second timeout.
The mean PAR-2 scores for \procount{} and \nest{} were 47 and 87.
Further comparison is needed in future work.


In some sense, projected model counting on Boolean formulas is a dual problem of \emph{maximum a posteriori hypothesis (MAP)} inference \cite{murphy2012machine,maua2015complexity,xue2016solving} on Bayesian networks \cite{pearl1985bayesian}: a projected model count has the form $\sum_X \max_Y f(X, Y)$, while a MAP probability has the form $\max_Y \sum_X f(X, Y)$.
% The decision version of MAP is NP\textsuperscript{PP}-complete \cite{park2004complexity}.
Both problems can be solved using variable elimination, but an elimination order may not freely interleave $X$ variables with $Y$ variables.
A valid variable order induces an \emph{evaluation tree} (similar to a project-join tree%
% in which each internal node projects out exactly one variable%
) \cite{park2004complexity}.
% The time complexity of exact algorithms is exponential in the width of a valid variable order.
As mentioned in \cite{park2004complexity}, exact MAP algorithms construct evaluation trees using constraint-satisfaction heuristics (similar to our planner \htb).
Our work goes further by constructing low-width graded project-join trees using tree-decomposition techniques (with our planner \Lg) and by performing efficient computations using compact ADDs (with our executor \Dmc).
% To adapt our framework for MAP, we can simply swap $\Sigma$-variables with $\exists$-variables. 
