\section{Discrete Integration}
\label{sec:wmc}

In discrete integration (also called constrained counting) the task is to count the total weight, subject to a given weight function, of the set of solutions of input constraints \cite{GSS08}. We represent weight functions as pseudo-Boolean functions. As is standard, we focus on literal-weight functions: 
\begin{definition}[Literal-Weight Function]
\label{def:literal-weight}
A pseudo-Boolean function $W: 2^X \rightarrow \mathbb{R}$ is \emph{literal-weight} function if there exist pseudo-Boolean functions $W_x: 2^{\{x\}} \rightarrow \mathbb{R}$ for all $x \in X$ such that $W = \prod_{x \in X} W_x$. 
\end{definition}
Literal-weight functions are also called \emph{log-linear} functions and can be employed to capture a wide variety of probabilistic distributions that arise from graphical models, conditional random fields, skip-gram models, and the like~\cite{KF09}.
$W_x(\{x\})$ indicates the weight of the positive literal $x$ in $W$ and $W_x(\emptyset)$ indicates the weight of the negative literal $\neg x$ in $W$.

This thesis primarily focuses on two subclasses of discrete integration: model counting and projected model counting.

\subsection{Model Counting}
In (weighted) model counting, the input constraints are given as a propositional formula, typically in CNF. Formally:
\begin{definition}[Weighted Model Count]
  Let $\varphi$ be a formula over Boolean variables $X$ and let $W: 2^X \rightarrow \mathbb{R}$ be a pseudo-Boolean function. We say that $(X, \phi, W)$ is an instance of \emph{weighted model counting}. The \emph{$W$-weighted model count} of $\varphi$ is
  $$W(\varphi) \equiv \sum_{\tau \in 2^X} \varphi(\tau) \cdot W(\tau).$$
\end{definition}

The \emph{unweighted model count} of a Boolean formula $\varphi$ is the number of solutions of $\varphi$, i.e. the $W$-weighted model count of $\varphi$ where $W(\tau) = 1$ for all $\tau \in 2^X$.


We focus in this work on weighted model counting, as opposed to \emph{unweighted model counting} where the weight function $W$ is constant. There are a variety of counters \cite{CW16,FHMW17,Thurley2006} that can perform only unweighted model counting and so we do not compare against them. Of particular note here is \tool{countAntom} \cite{BSB15}, a multi-core unweighted model counter. An interesting direction for future work is to explore the integration of weights into \tool{countAntom} and compare with tensor-network-based approaches to weighted model counting.

Existing approaches to weighted model counting can be split broadly into three categories: \emph{direct reasoning}, \emph{knowledge compilation}, and \emph{dynamic programming}. In counters based on direct reasoning (e.g., \tool{cachet} \cite{SBK05}), the idea is to reason directly about the CNF representation of $\varphi$. In counters based on knowledge compilation (e.g. \tool{miniC2D} \cite{OD15} and \tool{d4} \cite{LM17}), the idea is to compile $\varphi$ into an alternative representation on which counting is easy. In counters based on dynamic programming (e.g. \tool{ADDMC} \cite{DPV20} and \tool{gpuSAT2} \cite{FHWZ18,FHZ19}), the idea is to traverse the clause structure of $\varphi$. Tensor-network approaches to counting (e.g. \tool{TensorOrder} \cite{DDV19} and this work) are also based on dynamic programming. Dynamic programming approaches often utilize graph decompositions, which we define in the next section. 

Moreover, this thesis focuses on \emph{exact} discrete integration, where the exact total weight must be produced, as opposed to \emph{approximate} discrete integration where approximations (possibly with theoretical guarantees) are allowed in the result. %TODO: cite

\subsection{Projected Model Counting}
In (weighted) projected model counting, the input constraints are given as a propositional formula, typically in CNF, and a set of irrelevant, existential variables.
Formally:
\begin{definition}[Weighted Projected Model Count]
	Let $\phi$ be a Boolean formula, $\{X, Y\}$ be a partition of $\vars(\phi)$, and $W: 2^X \to \R$ be a pseudo-Boolean function. We say that $(X, Y, \phi, W)$ is an instance of \emph{weighted projected model counting}.
	The \emph{$W$-weighted $Y$-projected model count} of $\phi$ is
	$$\func{WPMC}(\phi, W, Y) \equiv \sum_{\tau \in 2^X} \pars{ W(\tau) \mult \max_{\sigma \in 2^Y} \phi(\tau \cup \sigma) }$$.
\end{definition}

Variables in $X$ are called \emph{relevant} or \emph{additive}, and variables in $Y$ are called \emph{irrelevant} or \emph{disjunctive}. 
Notice that model counting is a special case of projected model counting where all variables are relevant. Thus $W(\varphi) = \func{WPMC}(\phi, W, \emptyset)$ for all Boolean formulas $\varphi$ and weight functions $W: 2^{\vars(\varphi)} \rightarrow \mathbb{R}$. 

There are several recent tools for weighted projected model counting.
For example, \dfp{} uses decision decomposable negation normal form \cite{lagniez2019recursive}, \projmc{} leverages disjunctive decomposition \cite{lagniez2019recursive}, and \ssat{} combines counting with SAT techniques \cite{lee2017solving}. 
While our focus in Chapter \ref{ch:procount} is on deterministic exact weighted projected model counting, it is worth mentioning that various relaxations have also been studied, e.g., probabilistic \cite{sharma2019ganak}, approximate \cite{ermon2013taming,fremont2017maximum,soos2019bird}, or unweighted \cite{zawadzki2013generalization,aziz2015projected,mohle2018dualizing,hecher2020taming} projected model counting.